I/O simulation model with Linux page cache, integration and evaluation in SimGrid framework

### 2020-12-07 paper_io_sim.tex
edits in methods

@@

    \paragraph*{Flushing}
    The main simulated thread in the Memory Manager can flush the
    cache to write dirty data to disk. The data flushing simulation
    function takes the amount of data to flush as parameter. While
    the amount of data to flush is not reached and there are dirty
    blocks remaining in cache, this function traverses the sorted
    inactive list, then the sorted active list, and writes the
    least recently used dirty block to disk having set its dirty
    flag to 0. In case the amount of data to flush requires that a
    block be partially flushed, the block is split in two blocks,
    one that is flushed and one that remains dirty. Flushing time
    to disk is simulated with the storage model.

    \paragraph*{Cache eviction}
    Similarly to data flushing, our cache eviction simulation runs in
    the main thread. It frees up the page cache by traversing and deleting 
    least recently used clean data blocks in the inactive list.
    The amount of data to evict is passed as a parameter and data blocks are deleted 
    from the inactive list until total evicted data reaches the required amount,
    or until there is no clean block left in the list.
    If the last evicted block does not have to be entirely evicted, the block is split in two blocks,
    and only one of them is evicted.
    The cache eviction simulation does not contribute to the simulated time 
    since cache eviction time is negligible in real systems \tristan{a reference would be welcome}.

@@

    \paragraph*{Periodical flushing}
    Periodical flushing is simulated in the Memory Manager
    background thread. As in the Linux kernel, a dirty block
    in our model is considered expired if (1) it is dirty, and (2)
    the duration since its entry time is longer than a
    predefined expiration time. 
    Periodical flushing is simulated as an infinite loop in which 
    the Memory Manager searches and flushes these expired dirty blocks to disk (Algorithm~\ref{alg:pdflush}). 
    % From Tristan: the algorithm is quite straightforward, I don't think this is 
    % necessary
    % In each repetition, Memory Manager finds expired dirty blocks in two 
    % page cache LRU lists (line 8), simulates writes of data of these blocks 
    % to disk (line 9), and mark them as clean (line 10).
    % If the flushing time does not exceed our time interval, the thread is put 
    % to sleep for the remaining time (lines 11-13). 
    % Then based on the state of the host at current simulated time, 
    % the algorithm continues or finishes the loop.
    Because periodical flushing is simulated as a background thread, it can happen concurrently
    with disk I/O initiated by the main thread. This parallelism is handled by the 
    storage model and reflected in simulated I/O time.

@@

    \begin{algorithm}\caption{File chunk read simulation of IOController}
    \label{alg:read}
        \small
        \begin{algorithmic}[1]
            \Input
                \Desc{cs}{chunk size}
                \Desc{fn}{file name}
                \Desc{fs}{file size (assume to fit in memory)}
                \Desc{mm}{MemoryManager object}
                \Desc{sm}{storage simulation model}
               \EndInput
               \State disk\_read = min(cs, fs - mm.cached(fn))
               \State cache\_read = cs - disk\_read
            \State mm.flush(cs + disk\_read - mm.free\_mem - mm.evictable, fn) 
            \State mm.evict(cs + disk\_read - mm.free\_mem, fn) 
            \If {disk\_read $>$ 0}  \Comment{Read uncached data}    
                \State sm.read(disk\_read)  
                \State mm.add\_to\_cache(disk\_read, fn)     
            \EndIf
            \If {cache\_read $>$ 0} \Comment{Read cached}
                \State mm.cache\_read(cache\_read)  
            \EndIf
        \end{algorithmic}
    \end{algorithm}            
    As mentioned previously, our model reads and writes file chunks in a round-robin 
    fashion.
    When a simulated application reads a file chunk, it sends a chunk read request 
    to IOController.             
    The algorithm to simulate a chunk read implemented in IOController 
    is described in Algorithm~\ref{alg:read}.   
    First, we calculate the amount of uncached data that needs to be read 
    from disk, and the remaining amount is read from cache (line 7-8).
    We assume that after a file is read, there are two copies of the file in memory 
    (one in cache, one in anonymous memory). 
    Unfortunately, SimGrid lacks of application memory management and WRENCH 
    only use the memory capacity to check if the applications can be run. 
    Thus, in our model, when a chunk is read, one copy of the chunk 
    occupies anonymous memory, and the uncached part of that chunk is 
    read into cache.    
    The amount of required memory to read the chunk is calculated and 
    if there is not enough memory available, Memory Manager is called to 
    flush dirty data to disk and simulate flushing time with memory model (line 9). 
    This flushing is complemented by eviction in the Memory Manager (line 10). 
    Then, if there is uncached data, a disk read is simulated with disk model (line 12), 
    and this data of the file is added to cache (line 13).
    Finally, if cached data is read, Memory Manager is called to simulate a cache read  
    and update the corresponding data blocks in cache (line 16).
    \begin{algorithm}\caption{File chunk write simulation of IOController}
    \label{alg:write}
        \small
        \begin{algorithmic}[1]
            \Input
                \Desc{cs}{chunk size}
                \Desc{fn}{file name}
                \Desc{mm}{MemoryManager object}
                \Desc{sm}{storage simulation model}
               \EndInput
            \State remain\_dirty = dirty\_ratio * mm.avail\_mem - mm.dirty
            \If {remain\_dirty $>$ 0} \Comment{Write with memory bandwidth}
                \State mm.evict(min(cs, remain\_dirty) - mm.free\_mem)
                \State mem\_amt = min(cs, mm.free\_mem)
                \State mm.write(fn, mem\_amt) 
            \EndIf
            \If {mem\_amt $<$ cs}  \Comment{Write with disk bandwidth}
                \State mm.flush(cs - mem\_amt)  
                \State mm.evict(cs - mem\_amt  - mm.free\_mem) 
                \State mm.add\_to\_cache(fn, min(cs - mem\_amt, mm.free\_mem))
                \State sm.write(cs - mem\_amt)
            \EndIf

        \end{algorithmic}
    \end{algorithm}
    Algorithm~\ref{alg:write} describes our simulation of chunk write in 
    the IOController. 
    In the chunk write, data can be written to cache with memory bandwidth 
    without waiting for flushing or being throttled before the amount of 
    dirty data reaches dirty\_ratio, or otherwise written with disk bandwidth.
    Thus, our algorithm initially checks the remaining amount of data that 
    can be written as dirty data (line 6).
    If this amount is greater than 0, MemoryManager is asked to evict 
    data from cache to accommodate the written data (line 8). 
    If the amount to evict calculated and passed in the evict function is less than 0, 
    the eviction does nothing.
    After evicting cache, the actual amount of data that can be written to 
    page cache is calculated (line 9), a cache write is then simulated 
    with this amount through memory model and data written is added to 
    page cache LRU lists (line 10).
    After dirty\_ratio is reached, dirty data has to be flushed before new data 
    can be written to cache again. 
    If there is remaining chunk data that has not been written, we flush and 
    evict data from cache as much as possible so that the this data can be 
    written to cache (line 13-14). 
    Here, MemoryManager does not guarantee that the amount of flushed 
    and evicted data is enough for the write of remaining data. 
    In the case it is less than the required amount, the remaining data 
    being written to cache can be flushed and evicted right away. 
    Thus, the amount of remaining data written and kept in cache 
    after the write is limited by the amount free memory, and this amount 
    is added to cache (line 15). 
    Finally, the disk write is simulated with the remaining data amount (line 16), 
    and the chunk write algorithm finishes.

@@

            Since bandwidth sharing was not simulated, this prototype does not support
            concurrency: it is limited to single-threaded pipelines running on systems
            with a single-core CPU. We used it in a first validation of our simulation
            model, against a real sequential pipeline running on a real system.
            The source code of our Python prototype is available at
            \url{https://github.com/big-data-lab-team/paper-io-simulation/tree/master/exp/pysim}

            Having the model validated, we created simulators for different use cases
            using WRENCH.
            We implemented our model separately from existing WRENCH components,
            exposed an API for users, and added an command line argument to enable
            the model in WRENCH.
            This helps us to avoid unexpected impacts on WRENCH and
            independently update the model in the future.
            We then compared the results of the WRENCH, WRENCH-cache simulators
            the results of real execution on a real system.

            In this work, we use Python 3.7 to implement the single-threaded simulator,
            SimGrid 3.25 and WRENCH 1.6 for more multi-threaded simulators.
            SimGrid source code is available at \url{https://framagit.org/simgrid/simgrid},
            and WRENCH is at
            \url{https://github.com/wrench-project/wrench/tree/master/src/wrench/services/memory}.

@@

            The goal of our experiments is to evaluate our page cache
            simulation in single-threaded and multi-threaded applications
            accessing data on local and network file systems.
            We use a pipeline of sequential tasks where each task reads an input file,
            increments every byte of the input to emulate real processing, and writes
            the result to disk.
            The output of the previous task is the input of the next task.
            We also use a real application pipeline \textcolor{red}{[Maybe briefly describe this application's purpose (e.g. a standard fMRI preprocessing piepeline)]},
            to evaluate the applicability of our simulation model.

>>>>

% Fix link colors
\hypersetup{
    colorlinks = true,
    linkcolor=red,
    citecolor=red,
    urlcolor=blue,
    linktocpage % so that page numbers are clickable in toc
}

@@

    The main simulated thread in the Memory Manager can flush the
    cache to write dirty data to disk. The data flushing simulation
    function takes the amount of data to flush as parameter. While
    the amount of data to flush is not reached and there are dirty
    blocks remaining in cache, this function traverses the sorted
    inactive list, then the sorted active list, and writes the
    least recently used dirty block to disk having set its dirty
    flag to 0. In case the amount of data to flush requires that a
    block be partially flushed, the block is split in two blocks,
    one that is flushed and one that remains dirty. Flushing time
    to disk is simulated with the storage model.

    Similarly to data flushing, our cache eviction simulation runs in
    the main thread. It frees up the page cache by traversing and deleting 
    least recently used clean data blocks in the inactive list.
    The amount of data to evict is passed as a parameter and data blocks are deleted 
    from the inactive list until total evicted data reaches the required amount,
    or until there is no clean block left in the list.
    If the last evicted block does not have to be entirely evicted, the block is split in two blocks,
    and only one of them is evicted.
    The cache eviction simulation does not contribute to the simulated time 
    since cache eviction time is negligible in real systems \tristan{a reference would be welcome}.

@@

    Periodical flushing is simulated in the Memory Manager
    background thread. As in the Linux kernel, a dirty block
    in our model is considered expired if (1) it is dirty, and (2)
    the duration since its entry time is longer than a
    predefined expiration time. 
    Periodical flushing is simulated as an infinite loop in which 
    the Memory Manager searches and flushes these expired dirty blocks to disk (Algorithm~\ref{alg:pdflush}). 
    % From Tristan: the algorithm is quite straightforward, I don't think this is 
    % necessary
    % In each repetition, Memory Manager finds expired dirty blocks in two 
    % page cache LRU lists (line 8), simulates writes of data of these blocks 
    % to disk (line 9), and mark them as clean (line 10).
    % If the flushing time does not exceed our time interval, the thread is put 
    % to sleep for the remaining time (lines 11-13). 
    % Then based on the state of the host at current simulated time, 
    % the algorithm continues or finishes the loop.
    Because periodical flushing is simulated as a background thread, it can happen concurrently
    with disk I/O initiated by the main thread. This parallelism is handled by the 
    storage model and reflected in simulated I/O time.

@@

    \begin{algorithm}\caption{File chunk read simulation of IOController}
    \label{alg:read}
        \small
        \begin{algorithmic}[1]
            \Input
                \Desc{cs}{chunk size}
                \Desc{fn}{file name}
                \Desc{fs}{file size (assumed to fit in memory)}
                \Desc{mm}{MemoryManager object}
                \Desc{sm}{storage simulation model}
               \EndInput
               \State disk\_read = min(cs, fs - mm.cached(fn))
               \State cache\_read = cs - disk\_read
            \State mm.flush(cs + disk\_read - mm.free\_mem - mm.evictable, fn) 
            \State mm.evict(cs + disk\_read - mm.free\_mem, fn) 
            \If {disk\_read $>$ 0}  \Comment{Read uncached data}    
                \State sm.read(disk\_read)  
                \State mm.add\_to\_cache(disk\_read, fn)     
            \EndIf
            \If {cache\_read $>$ 0} \Comment{Read cached}
                \State mm.cache\_read(cache\_read)  
            \EndIf
        \end{algorithmic}
    \end{algorithm}
    As mentioned previously, our model reads and writes file chunks in a
    round-robin fashion. To read a file chunk, simulated applications send
    chunk read requests to the IOController which processes them using
    Algorithm~\ref{alg:read}. First, we calculate the amount of uncached
    data that needs to be read from disk, and the remaining amount is read
    from cache (line 7-8). The amount of memory required to read the chunk
    is calculated, corresponding to a copy of the chunk in anonymous memory
    and a copy of the chunk in cache \tristan{Create a variable
    "required\_mem" in the algorithm to clarify}.
    \tristan{add a call to mm in the algorithm to increment anonymous memory. Mention in the experiments that anonymous memory is flushed after each task.}
    If there is not enough available memory, the Memory Manager is called
    to flush dirty data (line 9). If necessary, flushing is complemented by
    eviction (line 10). When called with negative arguments, functions
    flush and evict simply return and do not do anything. Then, if there is
    uncached data, a disk read is simulated with disk model (line 12)
    \tristan{this is not consistent with the new Figure 1 where disk reads
    are done by the mm}, and this data of the file is added to cache (line
    13).
    Finally, if cached data needs to be read, the Memory Manager is called
    to simulate a cache read  
    and update the corresponding data blocks accordingly (line 16).

    \begin{algorithm}\caption{File chunk write simulation of IOController}
    \label{alg:write}
        \small
        \begin{algorithmic}[1]
            \Input
                \Desc{cs}{chunk size}
                \Desc{fn}{file name}
                \Desc{mm}{MemoryManager object}
                \Desc{sm}{storage simulation model}
               \EndInput
            \State remain\_dirty = dirty\_ratio * mm.avail\_mem - mm.dirty
            \If {remain\_dirty $>$ 0} \Comment{Write with memory bandwidth}
                \State mm.evict(min(cs, remain\_dirty) - mm.free\_mem)
                \State mem\_amt = min(cs, mm.free\_mem)
                \State mm.write(fn, mem\_amt) 
            \EndIf
            \If {mem\_amt $<$ cs}  \Comment{Write with disk bandwidth}
                \State mm.flush(cs - mem\_amt)  
                \State mm.evict(cs - mem\_amt  - mm.free\_mem) 
                \State mm.add\_to\_cache(fn, min(cs - mem\_amt, mm.free\_mem)) \tristan{why is data written to cache with "addtocache" here and with "write" at line 10?}
                \State sm.write(cs - mem\_amt) \tristan{According to Fig 1 the IOController is not supposed to interact directly with sm}
            \EndIf

        \end{algorithmic}
    \end{algorithm}
    Algorithm~\ref{alg:write} describes our simulation of chunk writes in 
    the IOController. 
    % Data is written to cache until the amount of dirty data in the cache 
    % exceeds the dirty ratio times the amount of available memory. Data is then written to disk.
    Our algorithm initially checks the  amount of data that 
    can be written as dirty data given the dirty ratio (line 6).
    If this amount is greater than 0, the Memory Manager is requested to evict 
    data from cache if necessary (line 8) \tristan{the difference between available and free memory should be explained somewhere}.
    After eviction, the amount of data that can be written to 
    page cache is calculated (line 9), and a cache write is simulated (line 10).
    To write the remaining data, we flush and 
    evict data as possible so that this data can be 
    written to cache (line 13-14). 
    The Memory Manager does not guarantee that the amount of flushed 
    and evicted data is enough to write the remaining data. 
    In the case it is less than the required amount, the remaining data 
    being written to cache can be flushed and evicted right away. 
    Thus, the amount of remaining data written and kept in cache 
    after the write is limited by the amount free memory, and this amount 
    is added to cache (line 15) \tristan{the previous two sentences are very unclear, I don't understand them so I can't reformulate.}. 
    Finally, the disk write is simulated with the remaining data amount (line 16), 
    and the chunk write algorithm finishes \tristan{see comment in algorithm}.

@@

            Since bandwidth sharing was not simulated, this prototype does not support
            concurrency: it is limited to single-threaded applications running on systems
            with a single-core CPU. We used this prototype for a first validation of our simulation
            model against a real sequential application running on a real system.
            The source code this prototype is available at
            \url{https://github.com/big-data-lab-team/paper-io-simulation/tree/master/exp/pysim}. We used Python 3.7.

            We also implemented our model in WRENCH, separately from existing WRENCH components. 
            Page cache simulation is available through a user API and can be activated using a command-line argument.
            We used SimGrid 3.25 and WRENCH 1.6. SimGrid source code is available at \url{https://framagit.org/simgrid/simgrid},
            and WRENCH is at \url{https://github.com/wrench-project/wrench/tree/master/src/wrench/services/memory}.
            \tristan{which commit did you use?}

\tristan{anything else interesting to mention about the implementation?}

@@

        Our experiments compared real executions with the original WRENCH,
        our WRENCH-cache extension, and our prototype Python simulator. We
        evaluated our page cache simulation in single-threaded and
        multi-threaded applications accessing data on local and network
        file systems.

            We used a pipeline of three sequential tasks where each task reads the output of the previous task,
            increments every byte of the input to emulate real
            processing, and writes the resulting data to disk. The first file was provided as input. We also used
            a real application pipeline \textcolor{red}{[Maybe briefly
            describe this application's purpose (e.g. a standard fMRI
            preprocessing piepeline)]}, to evaluate the applicability of
            our simulation model.
